Rao-Blackwellized smoother for Conditionally Linear and Gaussian models

### 2016-11-13 TexArxiv/clgm.tex
Abstract and Introduction

\begin{abstract}
This paper focuses on Sequential Monte Carlo based approximations of smoothing distributions in conditionally linear and Gaussian state spaces. Only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps for these models. The algorithm proposed by  \cite{briers:doucet:maskell:2010} extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps. Other Rao Blackwellised Sequential Monte Carlo smoothers introduced in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} are based on the forward backward decomposition of the marginal smoothing distribution% In \cite{kim:1994}, a structural approximation of the model is used to simplify the backward pass while \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} use Rao Blackwellisation steps
 and use explicit marginalization of the linear states in both forward and backward passes. In this paper, we propose an overview of these algorithms and introduce  compared to an extension of the Two-filter based algorithm of \cite{briers:doucet:maskell:2010} with additional explicit Rao Blackwellised steps to improve its performance. The procedures are applied to commodity markets which are described using a two factor model based on the spot price and a convenience yield. Parameters are estimated using  an Expectation Maximization and the performance of the proposed procedure is assessed with Chicago Mercantile Exchange crude oil data.

@@

Conditionally Linear and Gaussian Models (CLGM) play an important role in many applications; see \cite{sarkka:2013} and the references therein for an up-to-date account. An important feature of these models is that, conditionally on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian, which implies that conditionally to the sequence of labels, the filtering and the smoothing distribution of the 'continuous' states $(Z_1,\dots,Z_n)$ can be solved explicitly. To exploit this specific structure, it has been suggested in the pioneering paper of \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to solve the filtering problem by combining Sequential Monte Carlo (SMC) to sample indicator variables with the Kalman filter to compute the conditional distribution of the 'continuous' states sequence $(Z_j)_{j=1}^n$.

>>>>

\begin{abstract}
This paper focuses on Sequential Monte Carlo based approximations of smoothing distributions in conditionally linear and Gaussian state spaces. Only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps for these models. The algorithm proposed by  \cite{briers:doucet:maskell:2010} extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps. Other Rao Blackwellised Sequential Monte Carlo smoothers introduced in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} are based on the forward backward decomposition of the marginal smoothing distribution % In \cite{kim:1994}, a structural approximation of the model is used to simplify the backward pass while \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} use Rao Blackwellisation steps
 and use explicit marginalization of the linear states in both forward and backward passes. In this paper, we propose an overview of these algorithms and introduce an additional Rao Blackwellisation step for all procedures which allows to sample new particles in the backward pass to improve the performance of the algorithm.  The procedures are applied to commodity markets which are described using a two factor model based on the spot price and a convenience yield. Parameters are estimated using  an Expectation Maximization and the performance of the proposed procedure is assessed with Chicago Mercantile Exchange crude oil data.

@@

CLGM play an important role in many applications; see \cite{sarkka:2013} and the references therein for an up-to-date account. A crucial feature of these models is that, conditional on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian, which implies that conditional on the sequence of labels, the filtering and the smoothing distribution of the 'continuous' states $(Z_1,\dots,Z_n)$ can be solved explicitly. To exploit this specific structure, it has been suggested in the pioneering paper of \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to solve the filtering problem by combining Sequential Monte Carlo (SMC) to sample indicator variables with the Kalman filter to compute the conditional distribution of the 'continuous' states sequence $(Z_j)_{j=1}^n$.
